COMPJ-2022-05-0325: Response to referees

### 2023-02-20 response-to-referees/OLD deleted lambda notes.tex
Fantastic corrections noticed by Konrad Hinsen

\documentclass{comjnl}
\usepackage{ragged2e}
\usepackage{amsmath}
\usepackage{amssymb}

%----
  \setlength\textheight{220mm}
%----

\input paper-seb-macros.tex
\input generated-info-for-main.tex
\input generated-page-lengths.tex

\long\def\ignore#1{}

%\linenumbers

\title{\mytitle} 
\author{Harold Thimbleby}
\shortauthors{Harold Thimbleby}
\affiliation{See Change Fellow in Digital Health, Swansea University}
\email{harold@thimbleby.net}

%\subject{Computational science (bioinformatics,~medical~computing,~etc)}

\keywords{Computational Science; Software Engineering; Reproducibility; Scientific scrutiny; Reproducible Analytic Pathway\raise 1ex\hbox{$\star$} (\italicRAPstar\hskip -1pt)}

\received{30 April 2022}
%\revised{--}

%\begin{tabular}{@{}ll}Current Address: &62 Cyncoed Road, Cardiff, CF23 5SH, Wales\\
%&\texttt{harold@thimbleby.net}\\
%&ORCID 0000-0003-2222-4243
%\end{tabular}

%\orcid{0000-0003-2222-4243}

\begin{document}

\begin{abstract}
\justifying
\noindent
While data, analysis, and the results of running scientific models (including AI and ML) are described in peer reviewed papers, the algorithms and code generating insights and results often avoid adequate scrutiny. If code is not correct, then the contributions of published papers relying on unchecked code will be suspect.

\hskip 1em This paper shows that scientists rarely assure the structure and quality of code they rely on, rarely make full code available for wider use or scrutiny, and rarely provide adequate documentation to understand or use their code reliably. This paper therefore justifies and proposes ways to mature the computational sciences:

%\emph{Problems:} Assumptions in scientific code are hard to scrutinize as they are rarely made explicit, even when code is made available. Both algorithms and code have bugs, unknown and accidental assumptions that have unwanted effects. Code is fallible, so any interpretation that relies on code is also fallible. When the code is not clearly structured and published with adequate documentation, the code cannot be usefully scrutinized. In turn, scientific claims cannot be properly scrutinized.

%\emph{Methodology:} From the perspective of Software Engineering, this paper critiques the quality (and accessibility) of coding in computational science, particularly as relied on in leading COVID-19 pandemic research driving international public health interventions. 
%The paper reports a pilot survey of peer reviewed computational modeling papers ($N=\the\dataN$) published in leading scientific journals.

%Code can be improved using Software Engineering. This paper argues for specific solutions:

%\emph{Results:} %Journals do not always have a code policy, and many have a relaxed approach to code disclosure.

%\noindent\emph{Solutions:}

\raggedright
\newdimen \mywidth \mywidth=\textwidth
\advance \mywidth by -3em

\def\indented#1{\vskip 1ex
\parshape 2 0em \textwidth 2em \mywidth 
\noindent 
\hbox to 2em{ #1.\hfill}\ignorespaces}

\indented 1
Professional Software Engineers can help and should be involved, particularly in critical research such as public health, climate, etc;

\indented 2
``Software Engineering Boards'' (analogous to Ethics or Institutional Review Boards) should be instigated and used;

\indented 3
Code, when used, should be considered an intrinsic part of any publication, and therefore should be formally reviewed by competent Software Engineers;

\indented 4
\begin{change}The Reproducible Analytic Pathway (RAP) methodology should be generalized to cover code and Software Engineering methodologies, in a generalization this paper introduces and terms \RAPstarp. Furthermore \RAPstar\ should be supported and encouraged in journal, conference, and funding body policies.
\end{change}

\vskip 1ex 
\noindent
The \supplement\ for this paper provides a summary of professional Software Engineering best practice relevant to scientific research and publication. It also includes suggestions for \RAPstar\ processes, and a pilot survey of code quality in leading peer-reviewed journals that corroborates the concerns of the paper.

\begin{quote}\def\seprule{\vskip 2ex\hrule\vskip 2ex}
\seprule

\def\zq{\setbox0=\hbox{``}\hskip -0\wd0\copy0}
\zq Science is what we understand well enough to explain to a computer.''\\
\hfill {\rm Donald E. Knuth in $A=B$ \cite{a=b}}\\

\zq I have to write to discover what I am doing.''\\
\hfill {\rm Flannery O'Connor, quoted in \emph{Write for your life\/} \cite{write-for-life}}

\zq Criticism is the mother of methodology.''\\
\hfill {\rm Robert P. Abelson in \emph{Statistics as Principled Argument\/} \cite{abelson}}

\seprule
\end{quote}
\end{abstract}
\maketitle

%\section*{Author summary}
%
%%\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{HT}}]
%Harold Thimbleby PhD, FRCP (Edinburgh), Hon.\ FRSA, Hon.\ FRCP is See Change Fellow in Digital Health at Swansea University, Wales. His research focuses on human error and computer system design, particularly for healthcare. 
%%In addition to over 340 peer reviewed and 188 invited publications, Harold has written several books, including \emph{Press On\/} (MIT Press, 2007), which was winner of the American Association of Publishers best book in computer science award.
%Harold won the British Computer Society Wilkes Medal. He is emeritus Gresham Professor of Geometry (a chair founded in 1597), and has been a Royal Society-Leverhulme Trust Senior Research Fellow and a Royal Society-Wolfson Research Merit Award holder. His latest book is  \emph{Fix IT: How to see and solve the problems of digital healthcare\/} \cite{fixit}.  See his web site, \texttt{www.harold.thimbleby.net}, for more details.

\section{Introduction}
\begin{change}
Unreliable, often unstated and unexplored, computational dependencies in science are widespread. Furthermore, code is rarely published in any useful form or professionally scrutinized, and therefore the code itself does not contribute to furthering reliable science, for instance through replication or reproduction. In short, the quality of much computational science (including AI and ML) is undermined because the code it relies on is rarely of adequate quality for the uses to which it is put.

This paper explores the nature and extent of these Software Engineering problems in published science. The paper suggests ways to improve the reliability of computational sciences.

Code should be developed and discussed in a sufficiently professional, rigorous, and recognizable way that is able to support clear presentation and scrutiny. Developing justifiably reliable code is the concern of the field of Software Engineering, which is discussed further below, as well as substantially in the paper's \supplement, which is an integral part of the suggested solutions and literature survey.

This paper makes an explicit analogy between reliable computation and reliable statistics, a field that has adopted clear standards for reliable use and presentation. The \supplement\ (section~\ref{supplementary-Spiegelhalter-section}) summarizes Spiegelhater's uncontroversial statistics probes and  --- just as it is routine for statistics, code and results from code (and the data it is run on) need to be discussed and presented in a way that properly assure belief in any claims derived from using them.

\begin{table*}
\begin{center}\normalsize
\begin{tabular}{|rl|} \hline
\the\numberOfJournals&Journals\\
\the\dataN&Papers:\\
\tabularJournalBreakdown
\the\countAuthors&Published authors\\
\the\totalPages&Published journal pages\\
July 2020&Sample month\\ \hline
\end{tabular}
\end{center}
\caption{Overview of peer-reviewed paper sample, broken down further in Table \ref{table-summary}.}
\label{table-overview}
\end{table*}

\begin{table*}
\begin{center}\normalsize
\input generated-summary-table.tex
\end{center}
\caption{Summary of a pilot survey of computational science papers. Full details of the survey, all data, and methodology are in the \supplement.}
\label{table-summary}
\end{table*}
\end{change}

The body of this paper reviews some specific epidemiological computational science and its avoidable problems. The \supplement\ further describes a broad pilot survey of papers covering a wide range of computational science published in leading journals, confirming that the issues raised in the body of the paper are widespread and not specific to science that may have been rushed because of relevance to the COVID-19 pandemic. The quality of the supporting code in both epidemiological science and in the survey was very weak (see a summary in Tables \ref{table-overview} \& \ref{table-summary}). No papers provided any evidence their code was adequately tested or rigorously developed; none used methodologies like RAP or \RAPstarp, and only one paper discussed any relevant Software Engineering methods (independent coding). Although in the sample 81\% of papers were published in journals that had code policies (which themselves are weak), 42\% of surveyed papers in those journals breached their code policies. One paper declared it had accessible code, but the relevant repository was and remains empty. Such results about code are comparable to surveys of data and data access \citeeg{machine-learning-reproducibility,no-raw-data,data-access}, and form part of the wider reproducibility crisis \citeeg{reproducibility-crisis}.

\begin{change}
\section{Background}
The discoveries and inventions of scientific technologies and instruments like microscopes, telescopes, and X-rays, drive and expand the sciences. There are fascinating periods as new ideas and science unify; for example, thermometers could not measure temperature in any meaningful way until the science was mature. For over a century, there was no agreement on definitions of temperature, how to calibrate thermometers, or what units they measured in. Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having a consensus in turn depended on reproducible thermometer measurements and a thorough understanding of the science, including all the confounding factors that were being misunderstood \cite{temperature}. %For example, if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the mercury volume also increases with temperature (but at a different rate), so the volume measurement is inaccurate. 
Science matured from no quantitative interest in temperature, through a complex process of hand-in-hand theory development and technical maturation, to having robust, off-the-shelf instruments that measure temperature in reliable, internationally standardized units.

Far more flexible and challenging than thermometers, computers are a unique, new technology. Understanding computers and integrating them into science is likely to take longer than intrumentalizing temperature did. Computation not only expands science's paradigms and supports new discoveries, but it can also \emph{do\/} new science --- and with AI, robotics, and other techniques, computation can do both real and speculative investigations autonomously, and better than humans.

``Computational science'' has come to mean a particular style of science, based on developing and using explicit computational models, but, really, \emph{all\/} of science is now computational. Computational science is not just restricted to specialized fields like computational chemistry, genomics, big data~\ldots\ in all fields of science, computation is used at every step, from note taking, sound and photographic recording, literature searches, analysis and statistics, correspondence with co-authors and editors, through to typesetting finished publications.

All areas of science are now being profoundly computerized, and we need to do it well.

\subsection{The statistics/computation analogy}\label{statistics-computation-analogy}
The central role of computational methods in science may be fruitfully compared to statistics, an established scientific tool --- and also, of course, like computer science, a substantial field of research in its own right.

Poor statistics is much easier to do than good statistics, and there are many examples of science being let down by na\"\i vely planned and poorly implemented statistics. Often scientists do not realize the limitations of their own statistical skills, so careful scientists generally work closely with professional statisticians. 
\end{change}

In good science, all statistics, methods and results are reported very carefully and in precise detail \cite{apa-statistics,Mariusz-reporting}, generally following strict journal or disciplinary guidelines. For example, a statistical claim might be summarized as follows:

\begin{quote}\raggedright
\setbox0=\hbox{``}
\hskip -\wd0\box0 Random intercept linear mixed model suggesting significant time by intervention-arm interaction effect. 
\ldots\ Bonferroni adjusted estimated mean difference between intervention-arms at 8-weeks 2.52 $( 95\%$ CI $0.78, 4.27, p = 0.0009)$. Between group effect size $d = 0.55$ $(95\%$ CI $0.32,$ $0.77)$.'' \cite{example-stats}
\end{quote}

This typical wording formally summarizes confidence intervals, $p$ levels, and so on, to present statistical results so the paper's claims can be seen to be complete, easy to interpret, and easy to scrutinize. It is a \emph{lingua franca}. It may look technical, but it is written in the standard and now widely accepted form for summarizing statistics --- it is a clear, rigorous, and readily interpreted way to express uncertainty in results. Moreover, behind any such brief paragraph is a substantial, rigorous, and appropriate statistical analysis.

Scientists write like this and conferences and journals require it because statistical claims need to be properly accountable and documented in a clear way. Authors, too, must avoid the easy temptation of being unjustifiably over-confident with their own results.

Spiegelhalter \cite{Spiegelhalter} says statistical information needs to be accessible, intelligible, assessable, and usable; he also suggests probing questions to help assess statistical quality (see \supplement\ section~\ref{supplementary-Spiegelhalter-section}). Results should not be uncritically accepted just because they are claimed. The skill and effort required to do statistics so it can be communicated clearly and correctly, as above, is not to be taken for granted; in fact, there is widespread concern about the poor quality of statistics in science \cite{paul-hci,pnas-stats}. While it is assumed that statistics should be peer reviewed, and that review will often lead to improvement, these critical papers show that reviewers and editors are often failing to pick up on poor statistics.

Scientists accept that statistics is a distinct, professional science, itself subject of research and continual improvement. Among other implications of the depth and progress of the field of statistics, undergraduate statistics options for general scientists are insufficient training for rigorous work in science --- their main value, perhaps, is to help scientists to understand the value of collaborating with specialist statisticians. Collaboration with statisticians is particularly important when new types of work are undertaken, where the statistical pitfalls have not already been well-explored.

Except in the most trivial of cases, all numbers and graphs, along with the statistics underlying them, will be generated by computer. Indeed, computers are now very widely used, not just to calculate statistics, but to run the models, do the data sampling and processing, to operate the sensors or surveys that generate the data, and to process it. Many papers now explore the contribution of AI and ML to their fields. The data --- including the databases and bibliographic sources --- and code to analyze it is all stored and manipulated on computers. Computers even help with the word processing and typesetting of the research.

In short, computers, data, and computer code are central to modern science, not just to the explicitly computational sciences. Some AI work is uncovering biases and ethical issues that were previously unrecognized, so computational sciences are not just routinely contributing to existing science but extending its reach and improving its quality.

However, using any code raises many critical questions: formats, backup, cyber-vulnerability, version control, integrity checking (e.g., managing human error), auditing, debugging and testing, and more.  Software code, like statistics, is also subject to unintentional bias \cite{Ben,se-bias}.  All these issues are non-trivial concerns requiring technical expertise to manage well. As with statistics, good answers to such ``technical'' issues makes the science that relies on them better.

A common oversight in scientific papers is to present a model, such as a set of differential equations, but omit how that model is transformed into code that generates the results the paper summarizes; if so, the code may have problems that cannot be identified as there is no specification to reference it to.

Failure to properly document and explain computer code undermines the scientific value of the models and the results they generate, in the same way as failure to properly articulate statistics undermines the value of any scientific claims. Indeed, as few papers use code that is as well-understood and as well-researched as standard statistical procedures (such as Bonferroni corrections), the scientific problems of poorly reported code are widespread.

We would not believe a statistical claim that was obtained from some \emph{ad hoc\/} analysis with a new fangled method devised just for one project --- instead, we demand statistics that is recognizable, even traditional, so we are assured we understand what has been done and how reliable results were obtained.

An interesting overlap with statistical and Software Engineering sloppiness concerns the many papers that  disclose as part of their methodology that they used a particular package, for example

\begin{quote}\raggedright
\setbox0=\hbox{``}
\hskip -\wd0\box0 Data analyses were performed using SAS 9.2 (SAS Institute, Cary, North Carolina, USA)'' \cite{sas9.2}
\end{quote}

but without giving any further details. The paper cited above describes one of its multivariate analyses as ``multiple correspondence analysis (MCA) and ascendant hierarchical clustering'' with no specific details --- reproducing the work would be non-trivial since the methodology would have to be reconstructed from scratch, even assuming the data was made available by the authors as the paper gives no details of the data used or how to obtain it. The problem is that the common practice of declaring using a named computational system (such as SAS in this case) does not help scrutiny, as such systems can do almost anything. \emph{How\/} those analyses might have been performed is not discussed, and one assumes it follows that the analyses could therefore not have been reviewed for basic scientific competence during the publication process.

A reviewer, if nobody else, needs to actually examine the statistical code used and its documentation to assess whether the analysis presented in the paper is appropriate and sufficiently reliable. Furthermore, if the analysis actually depended on using SAS version 9.2, and not \emph{any\/} general purpose statistical system, then it is problematic because it is not reproducible if it relies on idiosyncrasies in SAS\@. Of course, an author can disclose the idiosyncratic dependencies on, say, version 9.2; while this seems to be an onerous obligation, conversely it is arguable that if an author is unaware of code dependencies, then their science relying on them is equally unreliable.

It is recognized that to make critical claims, models need to be run under varying assumptions \cite{whitty}, yet somehow it is easy to overlook that the code that implements those models also needs to be carefully tested under varying assumptions to uncover and fix bugs and biases, and to uncover unknown dependencies. Indeed, the code may be poorly written (as this paper shows is very likely), so the results derived from the code simply may not be reliable. Being able to understand (at least in principle --- e.g., relying on reviewers to have checked it) the exact code used in implementing a model is critical to having confidence in the results that rely on it.

In normal scientific reporting (outside of teaching and assessing science) details of methodology are routinely glossed. A chemist does not say they cleaned their glassware. One might argue, then, that scientists need not discuss their code in any detail because they know how to program and their code is correct. This argument is mistaken, as this paper shows. Unfortunately, code is rarely considered a valuable part of the science to which it contributes and this creates a vicious cycle of ignoring code, leading to ignoring the critical --- and non-trivial --- role of correct code in science.

\subsection{The role of code \\ in science and scientific publication}
%Yet without code, models could not be run.

Models map theory and parameters to describe phenomena, typically to make predictions, or to test and refine the theory supporting the models. With the possible exception of theoretical research, all but the simplest models require computers to evaluate; indeed even theoretical mathematics is now routinely performed by computer systems.

Whereas the mathematical form of a model may be concise and readily explained, even a basic computational representation of a model can easily run to thousands of lines of code, and its parameters --- its data --- may also be extensive. The chance that a thousand lines of code, hand-written or otherwise, is error free is negligible, and therefore good practice demands that checks and constraints should be applied to improve its reliability. How to do this well is the concern of Software Engineering, and is discussed throughout this paper and in the \supplement.

While scientific research may rely on relatively easily-scrutinized mathematical models, or models that seem in principle easy to mathematize, the models that are run on computers to obtain the results published are sometimes not disclosed, and even when they are they are long, complex, inscrutable and (our survey shows) lack adequate documentation. Therefore the models are very likely to be unreliable \emph{in principle}.

If code is not well-documented, this is not only a problem for reviewers and scientists reading the research to understand the intention of the code, but it also causes problems for the original researchers themselves: how can they understand its thinking well enough (e.g., a few weeks or months later) to maintain it correctly if it has not been clearly documented? As a scientist pursues a research career building on their previous work, how can they be certain their work is reliable, and not merely converging to their  prejudices? Without proper documentation, including a reasoned case to assure that the approach taken is sound \cite{assurance-case}, how do researchers, let alone reviewers, know exactly what they are doing?

Without substantial documentation it is impossible to scrutinize code properly. Consider just the single line ``\texttt{y = k*exp(x)}'' where there can be \emph{no\/} concept of its correctness \emph{unless\/} there is also an explicitly stated relation between the code and the mathematical specifications. What does it mean? What does \texttt{k} mean --- is it a basic constant or the result of some previous complex calculation? Does the code mean what was intended? What are the assumptions on \texttt{k}, \texttt{x}, and \texttt{y}, and do they hold invariantly? Moreover, as code generally consists of thousands of such lines, with numerous inter-dependencies, plus calling on many complex libraries of support code, it is inevitable that the \emph{collective\/} meaning will be unknown. A good programer would (in the example here) at least check that \texttt{k} and \texttt{x} are in range and that \texttt{k*exp} was behaving as expected (e.g., in case of under- or overflow).

Without explicit links to the relevant models (typically mathematics, depending on the claims), it is impossible to reason whether any code is correct, and in turn it is impossible to scientifically scrutinize results obtained from using the code. Not providing code and documentation, providing partial code, or providing code without the associated reasoning is analogous to claiming ``statistical results are significant'' without any discussion of the relevant methods and statistical details that justify making such a claim. If such an unjustified pseudo-statistical claim was made in a scientific paper, a reviewer would be justified in asking whether a competent experiment had even been performed. It would be generous to ask the author to provide the missing details so the paper could be better reviewed on resubmission.

Contrary to the views expressed in the present paper, some authors assert that the purpose of code is to provide insight into models, rather than precise (generally numerical) analyses summarizing data or properties of the data \cite{assessing-quality}. If code is inadequate, the so-called ``insights'' it provides will be flawed, and flawed in unquantified and unknown ways. Indeed, none of the papers sampled (described in section~\ref{survey-section}) claimed their papers were using code for insight; all papers claimed, explicitly or implicitly, that their code outputs were integral to their peer reviewed results.

Clearly, like statistics, programming (coding) can be done poorly and reported poorly, or it can be done well and reported well --- and any mix between these extremes. The question is whether it matters, \emph{when\/} it matters, and, if so, when it does, \emph{what\/} can be done to \emph{appropriately\/} help improve the quality of code (and discussions about the code) in scientific work?

\subsection{The deceptive simplicity of code}\label{deceptive-simplicity-of-code}
It is a misconception that programming is easy and even children can do it \cite{fixit}. More correctly, toy programming is easy, but real programming is very difficult.

An analogy helps. Building houses is very easy --- indeed, many of us have built toy Lego houses. Obviously, though, a Lego house is not a \emph{real\/} house. It is not large enough or strong enough for safe human habitation! This point is obvious because we can see Lego houses, and everyone is familiar with building-block play; its real-world engineering limitations are too obvious to need stating.

In contrast to Lego, computer programs are generally invisible, and therefore the engineering problems within them are also invisible. The ``programming is easy'' clich\' e is deceptive --- programming appears easy \emph{because\/} professional standards of building software are ignored, because people cannot see the reasons why they are needed, and because --- like Lego --- toy programs can look good but be unreliable, difficult to use and even dangerous. Thinking programming is easy is like appreciating a child's Lego building because we are not worried about subsidence, load bearing, electric shock, fire risks, water ingress, or even planning regulations. These are professional engineering issues that Lego builders ignore. Certainly, even real building is much easier and faster when the technical details are ignored, as anyone who has experienced a cowboy builder can attest.

Unlike building (the Code of Hammurabi dates to around 1755{\sc bc}), programming is a very new discipline, and the problems of poor programming are not widely appreciated or embedded in our culture. Relevant professional standards are not enforced. Problems for the reliability of science arise when legitimate doodling and tweaking software drifts into claiming scientific results that do not have reliable engineering structures underpinning them, let alone the properly developed and documented archived code, to justify them. In many countries, there are laws that require all but the very simplest structures to be formally approved from plans and inspected as they are built, but who writes plans for software, who inspects scientific models as they care coded? Yet the consequences of building a shoddy garage have negligible impact compared to the consequences of writing shoddy code that informs national public health policies or climate change interventions.

Quite different, more formal, reasoning confirms the misconception about programming. Lambda calculus ($\lambda$-calculus) is one of the very simplest of programming notations \cite{lambda}; it is just made up of strings of the four symbols $\lambda.()$ plus other symbols that have no pre-defined meanings. It is hard to think of a programming language that could look simpler, although of course anything can be written out in binary as a ``trivial'' sequence of \texttt{1}s and \texttt{0}s, but at least lambda calculus has a visible structure.

@@

Now consider this brief lambda expression: $\lambda t. (\lambda x. t (x x)) (\lambda x. t (x x))$. This expression  (the paradoxical combinator) allows us to easily define recursive functions \emph{without\/} any recursion. Such ``simple'' lambda calculus expressions are clearly profound. Indeed, lambda calculus has a place in the history of logic because it was the source of the first undecidable logic problems (the Halting Problem being perhaps the most famous example). Lambda calculus thus proves that simple programs are \emph{not\/} simple; just because children can write programs does not mean anybody, including the children, can understand them. Indeed, ``programming is easy'' is contradicted by the Halting Problem, which shows that at least one trivial question about a program, whether it halts, is in general unanswerable.

Since programming appears to be so easy, developing code has low status in scientific practice. Developers of code are rarely acknowledged in scientific papers. The implicit reasoning is: if programming is easy, then its intellectual contribution to science is negligible, so it is not even worth citing it or acknowledging the technical contributors to it. Because it is easy, there is no need to work hard to make it correct --- anyway, it obviously works. While such views prevail, the vicious cycle is that the low status means software development is done casually, which reinforces the low status.

In reaction to this vicious cycle, there is a growing movement to cite code correctly \cite{cite-code}, because code \emph{is\/} important, particularly for reproduction, testing and extension of any scientific work.

Few journals editorial policies recognize that data and code are theoretically and in practice indistinguishable (see \supplement). Given that data and code are equivalent and interchangeable, it follows that publishing policies on data handling should also apply at least as strictly to code.

\subsection{The central role of code is ignored}\label{central-role-of-code}
Almost all scientific papers \emph{routinely\/} describe their experimental method, their data handling, and provide an overview of their analytic (usually statistical) methods. If they are theoretical papers, they will describe their mathematical models and any data that is used to confirm their models. Outside of pure computer science, scientific papers are almost entirely silent on the code they rely on and how it was developed --- in particular, how the code might have been protected from bugs, analogously to how appropriate experimental methods avoid or control for experimental error. The relative silence on code is curious as science now depends on correct code as much as it ever depended on controlled experiments. While scientists avoid or manage contaminated samples, they ought to avoid or manage code bugs which have the same confounding impact.

It is important that experiments and analysis are performed reliably and ethically. Because statistics, like code, is so readily susceptible to uncontrolled bias and error, there are many protocols and journal policies that enforce best practice, for example journals often require adherence to PRISMA (Preferred Reporting Items for Systematic Reviews and Meta-Analyses) \cite{prisma} for any paper performing a systematic review of the literature. Yet PRISMA, like many such policies, ignores the critical role of code, and ignores the Software Engineering principles that assure computation is reliable and reliably reported.

PRISMA ``was designed to help systematic reviewers transparently report why the review was done, what the authors did, and what they found,'' which is reasonable enough. PRISMA covers the review process carefully. For example, the authors should report the number of papers they included in their review. Perhaps $N=2000$. This number is then written into their draft paper, likely in several places. As the authors read and revise their paper, submit it, and respond to peer reviewers, it is very likely that the number of papers included in the survey changes, or other numbers and details may change. The authors now have a maintenance problem: where are the numbers that have changed, and what should they be changed to? Doing a search-and-replace, whether automated or by hand, is fraught with difficulties. What happens if $2000$ is used for some other purposes as well? What happens if some of the $2000$ values are written as $2,000$ or $2000.0$ and are not noticed? What happens if some $2000$ are year dates and are changed incorrectly? Then there are the Human Factors:  slips and errors will happen in this process anyway. Typos, slips during cut-and-paste, and other errors are common. Similar iterative revision cycles happen with any paper, not just with systematic reviews. Avoidable errors are problems across all of science.

If a computer model is involved in the process (as it generally would be for a systematic review) then the value of $N$ in this example could very easily be stored in a file where the paper typesetting process can access it. For example, if \LaTeX\ is the system of choice, the analysis could generate, say,

\begin{center}\texttt{$\backslash$newcommand\{$\backslash$N-papers-reviewed\}\{2000\}}\end{center}

so that when and wherever the authors write \texttt{$\backslash$N-papers-reviewed} in their paper's text, the typeset paper says 2000 or whatever the correct value happens to be at the time. If so, then \emph{whenever\/} the survey is updated, the value cited in the paper is automatically and \emph{correctly\/} updated without any further intervention from the author.

In general, not just PRISMA and not just numbers, but any data, text, graphs or tables, etc, can be reliably inserted into a paper automatically.

PRISMA says nothing about how to ensure the final results of a survey are correctly and reliably presented in a paper, despite this being one of PRISMA's explicit motivations. Such rules reinforce the fallacy that code is trivial and unimportant.

These issues with PRISMA may sound trivial, but they are a symptom of the relaxed attitude to code that undermines the reliability of all published computational science.

\subsection{Bugs, code and programming}\label{knowledge}
Critiques of data and model assumptions are increasingly common \cite{critiques,diagnosis-reviews} but program code is rarely mentioned. Yet data and program are formally equivalent (see \supplement, section \ref{on-code-data-publication}). Program code has as great an effect on results as the data; in fact, without code, the data would be uninterpreted and probably useless. Code, however, is harder to scrutinize, which means that errors in code have subtle, often unnoticed effects on results.

Almost all code contains ``magic numbers'' --- that is, data masquerading as code. This common practice ensures that published data is very rarely all of the data because it omits the magic numbers embedded in the code. Data is often ``hard coded'' in programs with no explicit representation. Such issues emphasize the need for repositories to require the inclusion of code so all data, including that embedded in the code, is actually available.

Bugs can be understood as discrepancies between what code ought to do and what it actually does. Many bugs cause erroneous results, but bugs may be ``fail safe'' by causing a program to crash so no incorrect result can be delivered. Contracts and assertions are essential defensive programming technique that block compilation or block execution with incorrect results; they turn bugs into safe termination, or, better, failure to compile. None of the computational science surveyed or cited in this paper includes any such basic techniques.

If code is not documented it cannot be clear what it is intended to do, so it is not possible to detect and eliminate bugs. Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things \cite{essence-of-software,fixit} --- they are bugs that were intended but were ideas based on mistaken ideas (students and inexperienced programmers make intentional bugs all the time). For instance, in numerical modeling, using an inappropriate method can introduce errors that are not ``bugs'' in the narrow sense of incorrectly implementing what was wanted, but are bugs in the wider sense of producing incorrect results (e.g., ill-conditioning) that is, what was intended was wrong.

Random numbers are widely used in computational science, for simulation or for randomizing experiments. Misuse of random numbers (e.g., using standard libraries without testing them) is a very common cause of bugs nai\"\i ve bugs \cite{knuth}.

\begin{change}
\subsection{The long-term problems of unreliable code}
Scientists explore and extend the boundaries of rigorous knowledge. Put briefly, the purpose of scientific experiments is to vary details to either test and specify the boundaries of theories, or to discover new phenomena that lead to theory revision.

If poor code, or poorly documented code, is made available with scientific papers, the code (or the absence of code) is a natural place to start replicating and varying experimental conditions,  including both data or code. However, if the starting point is not accurately known, whether due to bugs, obscure code, or because of poor documentation, then experimental variations will have an unknown effect. Theory will then be driven by artifacts of the code, not genuine phenomena.

In section \ref{section-pandemic-modeling}, below, an example is documented of a research code development process of at least 15 years' duration where the code was admitted to be completely undocumented, leaving details in just one author's head. None of the various related papers describe any controls over the drift of the science, or how independent researchers building on it might have been able to build with confidence rather than merely reproducing the same errors (if any).

Since the code in question was substantial and non-trivial, it is very unlikely that any constructive reproduction occurred outside the original laboratory and mindset; indeed, section \ref{reproducibility-concerns} describes how ``reproduction'' became trivialized because of community pressure to confirm the insights of this particular research.

Trying to constructively refute aspects of this research in the Popperian sense \cite{popper-conjectures-refutations} would have been impossible. For example, had the relevant papers published critical code invariants then scientists building on the research could have explored whether those invariants remained valid and, if so, under what assumptions. In fact, invariants are the theories of code, and deserve as high a prominence in published computational science as the domain theories the computational modeling itself is supporting investigating.

Since the symmetries in the scientific domains should be reflected in corresponding symmetries in well-designed code, this raises new possibilities of Formal Methods approaches (e.g., using automatic theorem provers) supporting advances in general science.
\end{change}

\section{State of the art \\ in computational modeling}
\label{section-pandemic-modeling}
A review of epidemic modeling \cite{science-review} says, ``we use the words `computational modelling' loosely,'' and then, curiously, the review discusses exclusively mathematical modeling, implying that for the authors, and for the peer reviewers, there is no role for code or computation as such. It appears that the new insights, advances, rigor, and problems that computers bring to research were not considered relevant.

A systematic review \cite{diagnosis-reviews} of published COVID models for individual diagnosis and prognosis in clinical care, including apps and online tools, noted the common failure to follow standard TRIPOD guidelines \cite{tripod}. The review \cite{diagnosis-reviews} ignored the mapping from models to their implementation, yet if code is unreliable, the model \emph{cannot\/} be reliably used, and cannot be reliably interpreted regardless of whether TRIPOD guidelines are followed. Indeed, TRIPOD guidelines ignore code completely.

It should be noted that flowcharts, which the review \cite{diagnosis-reviews} did consider, are programs intended for direct human use. Flowcharts, too, should be designed as carefully as code, for exactly the same reason as it is hard to program reliably --- flowcharts can present arbitrarily complex algorithms and are often sketchy and vague on details that are critical to reliable use.

A high-profile 2020 COVID-19 model \cite{nature-summary,ICmodel} uses a modified 2005 computer program \cite{avianFluModel,originalICmodel} originally developed for modeling H5N1 in Thailand; it did not model air travel or other factors required for later western COVID-19 modeling. The 2020 model forms part of a series of papers \cite{ICmodel,avianFluModel,originalICmodel} none of which provide details of their code.

A co-author disclosed \cite{tweet} that the code was thousands of lines long and was undocumented C code. As Ferguson, the original code author, noted in an interview,

\begin{quote}\raggedright
\setbox0=\hbox{``}
\hskip -\wd0\box0 
For me the code is not a mess, but it's all in my head, completely undocumented. Nobody would be able to use it~\ldots
'' \cite{ferguson-interview}\end{quote}

This comment was made by a respected, influential world-leading scientist, with many peer-reviewed publications, and a respectable $h$-index\footnote{$h$-index: the largest value of $h$ such that at least $h$ papers by the author have each been cited at least $h$ times. The figure cited for Ferguson was obtained from Google Scholar on 20 January 2022. (Typical $h$ values vary by discipline.)} of 93. Ferguson should be well aware of the standards of coding used in at least his own field. This comment, quoted above, is arguably representative of the standards of the field as a whole.

Ferguson's admission is tantamount to saying that the published scientific findings are and need not be reproducible.\footnote{A constructive discussion of Software Engineering approaches to reproducibility can be found in \cite{basic-reproducibilty}.}

Lack of reproducibility is problematic, especially as the code would have required many non-trivial modifications to update it for COVID-19 with its different assumptions; moreover, the code would have had to have been updated very rapidly in response to the urgent COVID-19 crisis.

If Ferguson's C code had been made available for review, the reviewers would not have known how to evaluate it without the relevant documentation. It is, in fact, hard to imagine how a large undocumented program could have been repeatedly modified and repurposed over fifteen years without becoming incoherent, and someone recognizing that documenting it would help make it easier to maintain.

If code is undocumented, there would be an understandable temptation to modify it arbitrarily to get desired results; worse, without documentation and proper commenting, it is methodologically impossible to distinguish legitimate attempts at debugging from merely fudging the results. In contrast, if code is properly documented, the documentation defines the original intentions (including formally using mathematics to do so), and therefore any modifications will need to be justified and explained --- or the theory revised.

>>>>

\color{red}Quite different, more formal, reasoning confirms the misconception about programming. Lambda calculus ($\lambda$-calculus) is one of the very simplest of programming notations \cite{lambda}; it is just made up of strings of the four symbols $\lambda.()$ plus other symbols that have no pre-defined meanings. It is hard to think of a programming language that could look simpler, although of course anything can be written out in binary as a ``trivial'' sequence of \texttt{1}s and \texttt{0}s, but at least lambda calculus has a visible structure.

@@

Now consider this brief lambda expression: $\lambda t. (\lambda x. t (x x)) (\lambda x. t (x x))$. This expression  (the paradoxical combinator) allows us to easily define recursive functions \emph{without\/} any recursion. Such ``simple'' lambda calculus expressions are clearly profound. Indeed, lambda calculus has a place in the history of logic because it was the source of the first undecidable logic problems (the Halting Problem being perhaps a famous example). Lambda calculus thus proves that simple programs are \emph{not\/} simple; just because children can write programs does not mean anybody, including the children, can understand them. Indeed, ``programming is easy'' is contradicted by the Halting Problem, which shows that at least one trivial question about a program, whether it halts, is in general unanswerable.
\color{black}
